\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-02-26 analysis/manuscript.Rmd
Birgits suggestions

For over two decades, hybrid open access journal publishing, which makes some articles openly available while others remain behind a paywall, has been discussed as a means for transitioning the subscription system to full open access [@Prosser_2003].
The idea was that when journals increasingly publish open access articles, they could reduce revenues from subscriptions, while libraries and funders could change their funding models and shift expenditures from subscription to open access. 
However, initial approaches, mainly based on publication fees, also called article processing charges (APCs), did not contribute substantially to a large open access uptake. 
In 2009, Springer reported open access to 1% of articles in hybrid journals [@Dallmeier_2010].
Other studies also recorded a low uptake.
In 2011, only 1-2% of articles were open access [@Bj_rk_2012], growing to around 4% between 2011 and 2013 [@Laakso_2016].

With the introduction of central funding mechanisms for publication fees in some European countries since 2012, an increase in hybrid open access could be observed [@Bj_rk_2017; @Piwowar_2018; @Jubb_2017; @Huang_2020].
For example, studying university output, @Robinson_Garcia_2020 estimated a median uptake of 7.1% in the period 2014-2017. 
In particular, British (17%), Austrian (15%) and Dutch (13%) universities stand out. 
However, this shift in funding policy towards hybrid open access also added to the overall cost of publishing, which includes subscription spending and the administrative efforts required to handle payments [@Pinfield_2016].
Moreover, established large commercial publishers, which already dominated the publishing market [@Larivi_re_2015], disproportionately benefited from hybrid open access funding in comparison to full open access publishers [@Jahn_2016; @Butler_2023; @Shu_2023].

As a consequence, libraries and their consortia began to develop licensing strategies aimed at avoiding such ‘double dipping’ scenarios, in which well-established commercial publishers gain twice from reading and open access publishing [@Mittermaier_2015], as well as to increase publisher-provided immediate open access [@BJ_RK_2014; @Schimmer_2015].
These considerations resulted into transformative agreements[^agreements], which cover a broad range of contracts between library consortia and publishers from mid-2010s onwards where institutional spending for subscriptions and open access publishing are considered together [@Borrego_2020; @Hinchliffe_2019].
Transformative agreements seek to control costs while allowing a transitional phase for publishing more open access articles. 
Similar to big deals, transformative agreements mainly bundle hybrid and subscription-only journals from commercial publishers, but aim at a higher degree of transparency than previous big deals, where contracts including payments were confidential [@Bergstrom_2014].
<!-- For example, the nationwide DEAL consortium in Germany, which includes universities and academic institutions, collectively cancelled their subscriptions with Elsevier from 2017 onwards to negotiate a transformative agreement [@Fraser_2023]. -->

The introduction of transformative agreements aligns with funding policy changes, such as the decision made by the cOAlition S, a consortium of national funders including the European Commission, to no longer provide financial support for individual publication fees when publishing in hybrid journals.
According ot its Plan S launched in 2018, the cOAlition S funders only accept hybrid open access through transformative agreements "during a transition period that should be as short as possible" [@Schiltz_2018]. 
Specifically, they agreed to support hybrid open access only through transformative agreements from 2021, until the end of 2024.
Notably, the German Research Foundation (DFG), despite not being part of cOAlition S, has also extended its financial support for hybrid open access through transformative agreements [@Mittermaier_2021]. Previously, the DFG only provided funding for full open access journals [@Jahn_2016].

By the end of 2023, many transformative agreements were implemented, but interim outcomes are mixed.
The ESAC Transformative Agreement Registry[^esacregistry], the largest source of disclosure, recorded more than 800 transformative agreements, resulting in up to 900.000 open access articles published in both full open access and hybrid journals according to the accompanying ESAC Market Watch[^esacmarketwatch]. 
Library consortia reported increased open access volume, streamlined payment and monitoring procedures, as well as extensive utilization of open access options by the researchers they serve [@Marques_2020; @Pinhasi_2020; @Parmhed_2023]. 
The ongoing standardisation of transformative agreements contributed to improved transparency in terms of contracts and publisher-provided article metadata [@Marques_2019; @Pinhasi_2021].
However, with the growing trend toward transformative agreements, continued reliance on big deals is perceived as problematic, because it perpetuate market concentration [@Butler_2023; @Shu_2023]. 
Whether transformative agreements lead to reduced pricing remains uncertain [@Borrego_2023] and
a substantial transition of hybrid journals towards full open access could not be observed [@Momeni_2021; @Matthias_2019].
The focus on large commercial publishers might also increase inequality [@Klebel_2023], because transformative agreements focus on pay to publish open access mainly targets institutions from high-income countries, furthering a questionable journal prestige culture [@budapest]. 
Besides, an editorial-board resignation raised concerns that publishers' desire to maximize journal publication volume "without regard to quality" is a consequence of transformative agreements [@Goodin_2023].

The controversies surrounding hybrid open access and transformative agreements have led to varying policy conclusions. 
For instance, the Association of Swedish Higher Education Institutions (Sveriges universitets- och
högskoleförbund, SUHF) recommended to only support agreements for publishing in full open access journals [^sweden].
Likewise, most cOAlition S funders will discontinue financial support for transformative agreements by the end of 2024 [@Liverpool_2023].
The consortium also removed most hybrid journals from its Transformative Journal program in 2023 due to publishers' failure to meet self-defined open access growth targets [@Brainard_2023]. 
In contrast, the German DEAL consortium announced a five-year transformative agreement with Elsevier starting in 2024, while also renewing its contracts with Springer Nature and Wiley until the end of 2028. At the same time, the Colombia Consortium signed the first transformative agreements in Latin America [@Mu_oz_V_lez_2024].

Despite these controversies around transformative agreements as a means of transitioning journal publishing to full open access, there is limited evidence available on the adoption of open access in hybrid journals, and the extent to which this uptake can be attributed to transformative agreements.
Previous studies have focused on specific countries [@P_l_nen_2020; @Huang_2020; @Haucap_2021; @Wenaas_2022; @Taubert_2023] or publisher portfolios [@Pieper_2018; @Bakker_2024; @Fraser_2023; @Jahn_2021; @Momeni_2023], while large-scale studies relied on self-reported agreement data [@Moskovkin_2022], or used APC pricing lists [@Shu_2023]. 
Particularly, data availability is a limiting factor when studying the impact of transformative agreements [@Bakker_2024], because bibliometric databases, even though many allow the retrieval of open access articles in hybrid journals, do not directly attribute them to specific transformative agreements. 
Likewise, article-level open access invoicing data is only partly available [@Jahn_2021].

The present study aims to address these limitations by combining multiple openly available data sources to determine the open access uptake in hybrid journals, while distinguishing between open access through transformative agreements and other means. 
With this novel and open approach, this first large-scale analysis aims to answer the following questions:

@@

For both of these research questions, this study will analyse the variability by publisher, journal subject, and country.

@@

This study combines data from multiple publicly available data sources as shown in Figure \ref{fig:data_workflow}. 
Initially, transformative agreement data retrieved from the cOAlition S Journal Checker Tool[^checkertool]  provided information about journal portfolios and participating institutions. 
After identification of hybrid journals by excluding full open access journals, Crossref served as the primary data source for article-level metadata including Creative Commons (CC) license information to indicate open access availability on publisher websites.
To determine open access articles published through transformative agreements, first author affiliations from OpenAlex [@priem2022openalex] were subsequently linked to eligible institutions according to the transformative agreement data. 
In the following, these steps are described in more detail.

@@

A limitation of using the Journal Checker Tool and its underlying publicly available data dump to study the development of transformative agreements over time is that expired transformative agreements are constantly removed. 
To address this, four different snapshots were safeguarded and combined for this study: self-archived versions from July 2021, July 2022, and May 2023, as well as the most current dump downloaded on 11 December 2023.
This ensured that transformative agreements, which ended from 2021 onwards, were included, representing the majority of transformative agreements.
<!-- Agreements ending before 2021 could not be obtained.-->
Overall, the four combined Transformative Agreement Data dumps used in this study contained `r  length(unique(hoaddata::jct_hybrid_jns$esac_id))` out of `r length(unique(esac_df$ID))` agreements listed in the ESAC registry by December 2023.

The Transformative Agreement Data dumps link agreements to journals represented by journal names and ISSN. 
After mapping ISSN variants to the corresponding linking ISSN (ISSN-L) as provided by the ISSN International Centre, journals were associated to publishers using the ESAC ID, a unique identifier for transformative agreements in the ESAC Transformative Agreement Registry. Furthermore, journal subjects according to the All Science Journal Classification code (ASJC) were added from the Scopus journal source list as of August 2023.

Because transformative agreements can include both full open access and hybrid journals, the data were complemented with information about a journal's open access status using multiple sources: the Directory of Open Access Journals (DOAJ) downloaded on 12 December 2023[^doaj], OpenAlex (November 2023) and the the Bielefeld list of GOLD OA journals [@goldoa]. 
As shown in Figure \ref{fig:method_fig}A, combining different data sources considerably extended the journal matching<!-- to exclude full open access journals from transformative agreements-->.
In total, `r n_oa_journals` full open access journals were excluded based on ISSN matching.
The overlap between the three data sources was `r round(length(unique(Reduce(intersect, my_venn))) / length(unique(Reduce(union, my_venn)))* 100,0)`%.
The Gold OA journals dataset alone added `r length(unique(Reduce(setdiff, my_venn)))` journals, while the DOAJ comprised `r length(setdiff(my_venn$DOAJ, union(my_venn[["GOLD OA"]], my_venn$OpenAlex)))` full open access journals not listed in either of the other two sources. 
These full open access journals were mostly launched in 2022.

@@

After identifying hybrid journals included in transformative agreements, article metadata was retrieved from the Crossref <!-- and mapped to OpenAlex for obtaining author metadata including affiliations. --> November 2023 database snapshot for the five-year period 2018 to 2022 according to the issued date, representing the earliest known publication date. 
Because Crossref metadata lacked information to distinguish between original research articles including review and other types of journal content, which are often not covered by transformative agreements [@Borrego_2020], only articles published in regular issues indicated by non-numeric pagination were included.
Furthermore, an expanded version of Unpaywall’s paratext recognition approach was applied to exclude non-scholarly journal content such as table of contents.

Open access articles in hybrid journals were identified through Creative Commons (CC) license information in Crossref metadata. 
License information relative to the "accepted manuscript (AM)" version were not considered. 
Crossref was used for open access identification because transformative agreements workflows generally require publishers to deliver CC license information to this DOI registration agency [@Geschuhn_2017].

Comparing Crossref license coverage with OpenAlex, which re-uses open access evidence from Unpaywall, a widely used open access discovery service that also parses journal websites for open content licenses [@Piwowar_2018], highlighted ongoing challenges to identify hybrid open access [@Mart_n_Mart_n_2018; @Jahn_2021; @Zhang_2022; @Butler_2023].
Here, `r hoaddata::cr_upw |> filter(cr_year %in% 2018:2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, cr_hybrid_total) |> pull(cr_hybrid_total) |> sum()` articles with CC license were retrieved using Crossref, while `r hoaddata::cr_upw |> filter(cr_year %in% 2018:2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, upw_hybrid_total) |> pull(upw_hybrid_total) |> sum()` articles were tagged as "hybrid" according to the OpenAlex November 2023 release, which was used throughout this study.
The biggest differences concerned articles published between 2018 and 2020.
In 2022, however, Crossref and OpenAlex open access numbers only differ slightly (`r hoaddata::cr_upw |> filter(cr_year %in% 2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, cr_hybrid_total) |> pull(cr_hybrid_total) |> sum()` records using Crossref vs. `r hoaddata::cr_upw |> filter(cr_year %in% 2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, upw_hybrid_total) |> pull(upw_hybrid_total) |> sum()` in OpenAlex).
Notable difference could be furthermore observed among some publishers that presumably did not provide CC license information to Crossref including AIP Publishing, American Physiological Society, Emerald and the Royal Society.
Crossref license metadata was more complete with regard to articles from the publisher Wiley and American Chemical Society.
Finally, inconsistent open access status information in previous OpenAlex versions was observed [@jahn2023]. After reporting it to OpenAlex, fixing this issue was still ongoing according to the release notes, which might also explain part of the discrepancy.

After retrieving article metadata, the publication volume including open access was calculated per journal. 
To improve the identification of hybrid journals, journals with an open access proportion above 95% were excluded.
This further step allowed to remove additional `r length(unique(oa_prop$issn_l))` full open access journals.

Affiliation metadata about corresponding authors are crucial for the planning and evaluation of transformative agreements, because they are considered to be responsible to arrange open access publication [@Schimmer_2015; @Geschuhn_2017; @Borrego_2020]. 
Here, country and institutional affiliations were retrieved from OpenAlex.
However, because of low coverage in OpenAlex, this study focused on first authors and their affiliations instead.
First authors typically contribute most to a paper and are often considered lead author research papers [@Larivi_re_2016]. 
Related studies therefore assumed first authors as a proxy to measure to open access payments and the impact of transformative agreement [@Zhang_2022; @Shu_2023; @Haucap_2021].
Overall, around 90% of studied articles had first author affiliation metadata in OpenAlex, whereas the coverage of articles with corresponding author information was around 54%.

@@

To assess the impact of transformative agreements to hybrid open access,
participating institutions from the Transformative Agreement Data dump,
which were crowd-sourced from the agreements and consortia that
successfully negotiated an agreement, were matched with first author
affiliations recorded by OpenAlex using the ROR-ID.
The matching also took into account the duration of an agreement according to the ESAC registry.
Upon inspection, Transformative Agreement Data did not cover associated institutions comprehensively like university hospitals or institutes of large research organisations like the Max Planck Society. To improve the matching, Transformative Agreement Data was automatically enriched with associated organisations using OpenAlex's institution entity data.

In total, the compiled data set consists of `r hoaddata::jn_ind |> filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |> distinct(issn_l, cr_year, jn_all) |> pull(jn_all) |> sum()` articles published in `r hoaddata::jn_ind |> filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |> distinct(issn_l) |> nrow()` hybrid journals between 2018 and 2022 (see Figure \ref{fig:method_fig}B). 
Hybrid journals in transformative agreements represented `r cr_stats |> filter(src == "hoad") |> pull(n_articles) |> sum() / cr_stats |> pull(n_articles) |> sum() * 100`%  of total global output over the same time period according to Crossref, while full open access journals recorded 
`r cr_stats |> filter(is.na(src) & oa_status == "gold") |> pull(n_articles) |> sum() / cr_stats |> pull(n_articles) |> sum() * 100`%.

```{r method_fig, fig.asp=0.55, fig.cap="Initial data characteristics. (A) Full open access journals included in transformative agreements by evidence source Directory of Open Access Journals (DOAJ), OpenAlex and the Bielefeld GOLD OA list. (B) Number of articles in Crossref by journal types. The blue bars show the overall article volume of hybrid journals in transformative agreements, which were initially included in the study, in comaprision with full open access journals according to OpenAlex. The remainder represents closed access journals not covered by transformative agreements."}
(p | cr_stats_plot) + plot_annotation(tag_levels = 'A') +  plot_layout(widths = c(0.6, 0.4))
```
<!-- To reflect the increasing trend towards multiple and cross-country affiliations (Hottenrott et al., 2021; Larivière et al., 2015), articles were counted once per country. -->

@@

Throughout this mostly automated data gathering and analysis process, tools from the Tidyverse (Wickham et al., 2019) for the R programming language (R Core Team, 2020) were used. 
The resulting data is openly available through an R data package, hoaddata, version 0.2.91 
Following @Marwick_2018, hoaddata contains not only the datasets used in the data analysis. 
It also includes code used to compile the data by connecting it the cloud-based Google Big Query data warehouse, where the big scholarly data from Crossref, OpenAlex and Unpaywall were imported.
To increase the computational reproducibility, the data aggregation through hoaddata was automatically carried out using GitHub Actions, a continuous integration service.

@@

Between 2018 and 2022, a total of `r n_hybrid_jns` out of `r n_hybrid_jns_all` hybrid journals in transformative agreements published at least one open access article under a Creative Commons license.
During this period, these hybrid journals provided open access to `r n_oa` out of `r n_total` articles, representing a five-year open access proportion of `r n_oa / n_total * 100`%.
Authors who could make use of transformative agreements at the time of publication contributed `r n_ta_oa` open access articles to the total.

@@

Figure \ref{fig:results_overview}A shows a moderate growth in the proportion of open access articles in hybrid journals, comparing the overall open access uptake and the impact of transformative agreements on this trend.
Over the five-years period from 2018 to 2022, open access increased from `r oa_by_year |> filter(cr_year == "2018") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2018") |> pull(oa)`) to `r oa_by_year |> filter(cr_year == "2022") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2022") |> pull(oa)`). 
At the same time, the total article volume of the investigated journals grew from `r oa_by_year |> filter(cr_year == "2018") |> pull(articles)` in 2018 to `r oa_by_year |> filter(cr_year == "2022") |> pull(articles)` in 2022.

>>>>

For over two decades, hybrid open access journal publishing, which makes some articles openly available while others remain behind a paywall, has been discussed as a means for transitioning the subscription system to full open access [@Prosser_2003].
The idea was that when journals increasingly publish open access articles, they could reduce revenues from subscriptions, while libraries and funders could change their funding models and shift expenditures from subscription to open access. 
However, initial approaches, mainly based on publication fees, also called article processing charges (APCs), did not contribute substantially to a large open access uptake. 
In 2009, Springer reported open access to 1% of articles in hybrid journals [@Dallmeier_2010].
Other studies have also recorded low uptake.
In 2011, only 1-2% of articles were open access [@Bj_rk_2012], increasing to around 4% between 2011 and 2013 [@Laakso_2016].

With the introduction of central funding mechanisms for publication fees in some European countries since 2012, an substantial increase in hybrid open access has been observed [@Bj_rk_2017; @Piwowar_2018; @Jubb_2017; @Huang_2020].
For example, studying university output, @Robinson_Garcia_2020 estimated a median uptake of 7.1% in the period 2014-2017. 
In particular, British (17%), Austrian (15%) and Dutch (13%) universities stood out. 
However, this shift in funding policy towards hybrid open access also added to the overall cost of publishing, which includes subscription spending and the administrative efforts required to handle payments [@Pinfield_2016].
Moreover, large commercial publishers, which already dominated the publishing market [@Larivi_re_2015], disproportionately benefited from hybrid open access funding in comparison to full open access publishers [@Jahn_2016; @Butler_2023; @Shu_2023].

As a consequence, libraries and their consortia began to develop licensing strategies aimed at avoiding such ‘double dipping’ scenarios, in which well-established commercial publishers gain twice from reading and open access publishing [@Mittermaier_2015], as well as to increase publisher-provided immediate open access [@BJ_RK_2014; @Schimmer_2015].
These considerations resulted in transformative agreements[^agreements], which cover a broad range of contracts between library consortia and publishers from the mid-2010s onwards, where institutional spending for subscriptions and open access publishing are considered together [@Borrego_2020; @Hinchliffe_2019].
Transformative agreements seek to control costs while allowing a transitional phase for publishing more open access articles. 
Similar to big deals, transformative agreements mainly bundle hybrid and subscription-only journals from commercial publishers, but aim at a higher degree of transparency than previous big deals, where contracts including payments were confidential [@Bergstrom_2014].
<!-- For example, the nationwide DEAL consortium in Germany, which includes universities and academic institutions, collectively cancelled their subscriptions with Elsevier from 2017 onwards to negotiate a transformative agreement [@Fraser_2023]. -->

The introduction of transformative agreements aligns with funding policy changes, such as the decision made by cOAlition S, a consortium of funders and research organisations including the European Commission, to no longer provide financial support for individual publication fees when publishing in hybrid journals.
According to its Plan S launched in 2018, cOAlition S members only accept hybrid open access through transformative agreements "during a transition period that should be as short as possible" [@Schiltz_2018]. 
Specifically, they agreed to support hybrid open access only through transformative agreements from 2021, until the end of 2024.
Notably, despite not being part of cOAlition S, the German Research Foundation (DFG), has also extended its financial support for hybrid open access through transformative agreements [@Mittermaier_2021]. Previously, the DFG only provided funding for fully open access journals [@Jahn_2016].

By the end of 2023, many transformative agreements had been implemented, but the interim outcomes were mixed.
The ESAC Transformative Agreement Registry[^esacregistry], the largest source of disclosure, recorded more than 800 transformative agreements, resulting in up to 900.000 open access articles published in both fully open access and hybrid journals according to the accompanying ESAC Market Watch[^esacmarketwatch]. 
Library consortia reported increased open access volume, streamlined payment and monitoring procedures, as well as extensive utilisation of open access options by the researchers they serve [@Marques_2020; @Pinhasi_2020; @Parmhed_2023]. 
The ongoing standardisation of transformative agreements has contributed to improved transparency in terms of contracts and publisher-provided article metadata [@Marques_2019; @Pinhasi_2021].
However, with the growing trend toward transformative agreements, continued reliance on big deals is perceived as problematic, because it perpetuates market concentration [@Butler_2023; @Shu_2023]. 
Whether transformative agreements lead to reduced pricing remains uncertain [@Borrego_2023] and
a substantial transition of hybrid journals towards full open access could not be observed [@Momeni_2021; @Matthias_2019].
The focus on large commercial publishers might also increase inequality [@Klebel_2023], because transformative agreements' focus on pay to publish mainly targets institutions from high-income countries, furthering a questionable journal prestige culture [@budapest]. 
Besides, an editorial-board resignation raised concerns that transformative agreements might encourage publishers to maximize journal publication volume "without regard to quality" [@Goodin_2023].

The controversies surrounding hybrid open access and transformative agreements have led to varying policy conclusions. 
For instance, the Association of Swedish Higher Education Institutions (Sveriges universitets- och
högskoleförbund, SUHF) recommended only supporting agreements for publishing in fully open access journals [^sweden].
Likewise, most cOAlition S funders will discontinue financial support for transformative agreements by the end of 2024 [@Liverpool_2023].
The consortium also removed the majority of hybrid journals from its Transformative Journal program in 2023 due to publishers' failure to meet self-defined open access growth targets [@Brainard_2023]. 
By contrast, the German DEAL consortium announced a five-year transformative agreement with Elsevier starting in 2024 and also renewed its contracts with Springer Nature and Wiley until the end of 2028. Similarly, the Colombia Consortium signed the first transformative agreements in Latin America [@Mu_oz_V_lez_2024].

Despite these controversies around transformative agreements as a means of transitioning journal publishing to full open access, there is limited evidence available on the adoption of open access in hybrid journals, and the extent to which this can be attributed to transformative agreements.
Previous studies have focused on specific countries [@P_l_nen_2020; @Huang_2020; @Haucap_2021; @Wenaas_2022; @Taubert_2023] or publisher portfolios [@Pieper_2018; @Bakker_2024; @Fraser_2023; @Jahn_2021; @Momeni_2023], while large-scale studies relied on self-reported agreement data [@Moskovkin_2022], or used APC pricing lists [@Shu_2023]. 
In particular, data availability is a limiting factor when studying the impact of transformative agreements [@Bakker_2024], because bibliometric databases, even though many allow the retrieval of open access articles in hybrid journals, do not directly attribute them to specific transformative agreements. 
Likewise, article-level open access invoicing data are only partly available [@Jahn_2021].

The present study aims to address these limitations by combining multiple openly available data sources to determine open access uptake in hybrid journals, while distinguishing between open access through transformative agreements and other means. 
With this novel and open approach, this first large-scale analysis will answer the following questions:

@@

For both research questions, this study will analyse the variability by publisher, journal subject, and country.

@@

This study combines data from multiple publicly available sources, as shown in Figure \ref{fig:data_workflow}. 
Initially, transformative agreement data retrieved from cOAlition S Journal Checker Tool[^checkertool]  provided information about journal portfolios and participating institutions. 
After identifying hybrid journals by excluding fully open access journals, Crossref served as the primary data source for article-level metadata including Creative Commons (CC) license information to indicate open access availability on publisher websites.
To determine open access articles published through transformative agreements, first author affiliations from OpenAlex [@priem2022openalex] were subsequently linked to eligible institutions according to the transformative agreement data. 
In the following, the steps are described in detail.

@@

A limitation of using the Journal Checker Tool and its underlying publicly available data dump to study the development of transformative agreements over time is that expired transformative agreements are constantly removed. 
To address this, four different snapshots were safeguarded and combined for this study: self-archived versions from July 2021, July 2022, and May 2023, as well as the most current dump downloaded on 11 December 2023.
This ensured that transformative agreements, which ended from 2021 onwards, were included, representing the majority of transformative agreements.
<!-- Agreements ending before 2021 could not be obtained.-->
Overall, the four combined Transformative Agreement Data dumps used in this study contained `r  length(unique(hoaddata::jct_hybrid_jns$esac_id))` of `r length(unique(esac_df$ID))` agreements listed in the ESAC registry by December 2023.

The Transformative Agreement Data dumps link agreements to journals represented by journal names and the ISSN. 
After mapping ISSN variants to the corresponding linking ISSN (ISSN-L), as provided by the ISSN International Centre, journals were associated with publishers using the ESAC ID, a unique identifier for transformative agreements in the ESAC Transformative Agreement Registry. Furthermore, journal subjects according to the All Science Journal Classification code (ASJC) were added from the Scopus journal source list as of August 2023.

Because transformative agreements can include both fully open access and hybrid journals, the data were complemented with information about a journal's open access status using multiple sources: the Directory of Open Access Journals (DOAJ) downloaded on 12 December 2023[^doaj], OpenAlex (November 2023) and the Bielefeld list of GOLD OA journals [@goldoa]. 
As shown in Figure \ref{fig:method_fig}A, combining different data sources considerably extended journal matching<!-- to exclude full open access journals from transformative agreements-->.
In total, `r n_oa_journals` full open access journals were excluded based on ISSN matching.
The overlap between the three data sources was `r round(length(unique(Reduce(intersect, my_venn))) / length(unique(Reduce(union, my_venn)))* 100,0)`%.
The Gold OA journals dataset alone added `r length(unique(Reduce(setdiff, my_venn)))` journals, while the DOAJ comprised `r length(setdiff(my_venn$DOAJ, union(my_venn[["GOLD OA"]], my_venn$OpenAlex)))` fully open access journals not listed in either of the other two sources. 
These fully open access journals were launched in 2022.

@@

After identifying hybrid journals included in transformative agreements, article metadata were retrieved from the Crossref <!-- and mapped to OpenAlex for obtaining author metadata including affiliations. --> November 2023 database snapshot for the five-year period from 2018 to 2022, according to the issued date, representing the earliest known publication date. 
Because Crossref metadata lacked information to distinguish between original research articles, including reviews, and other types of journal content, which are often not covered by transformative agreements [@Borrego_2020], only articles published in regular issues indicated by non-numeric pagination were included.
Furthermore, an expanded version of Unpaywall’s paratext recognition approach was applied to exclude non-scholarly journal content such as table of contents.

Open access articles in hybrid journals were identified using the Creative Commons (CC) license information in Crossref metadata. 
License information relative to the "accepted manuscript (AM)" version was not considered. 
Crossref was used for open access identification because transformative agreement workflows generally require publishers to deliver CC license information to this DOI registration agency [@Geschuhn_2017].

Comparing Crossref license coverage with OpenAlex, which re-uses open access evidence from Unpaywall, a widely used open access discovery service that also parses journal websites for open content licenses [@Piwowar_2018], highlighted ongoing challenges in identifying hybrid open access [@Mart_n_Mart_n_2018; @Jahn_2021; @Zhang_2022; @Butler_2023].
Here, `r hoaddata::cr_upw |> filter(cr_year %in% 2018:2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, cr_hybrid_total) |> pull(cr_hybrid_total) |> sum()` articles under CC license were retrieved using Crossref, while `r hoaddata::cr_upw |> filter(cr_year %in% 2018:2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, upw_hybrid_total) |> pull(upw_hybrid_total) |> sum()` articles were tagged as "hybrid" according to the OpenAlex November 2023 release, which was used throughout this study.
The largest differences concerned articles published between 2018 and 2020.
In 2022, however, Crossref and OpenAlex open access numbers differ only slightly (`r hoaddata::cr_upw |> filter(cr_year %in% 2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, cr_hybrid_total) |> pull(cr_hybrid_total) |> sum()` records using Crossref vs. `r hoaddata::cr_upw |> filter(cr_year %in% 2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, upw_hybrid_total) |> pull(upw_hybrid_total) |> sum()` in OpenAlex).
Notable differences could be observed among some publishers that presumably did not provide CC license information to Crossref, including AIP Publishing, the American Physiological Society, Emerald, and the Royal Society.
Crossref license metadata was more complete with regard to the articles published by Wiley and the American Chemical Society.
Finally, inconsistent open access status information in previous OpenAlex versions was observed [@jahn2023]. After reporting this to OpenAlex, fixing this issue was still ongoing according to the release notes, which might also explain part of the discrepancy.

After retrieving the article metadata, the publication volume, including open access, was calculated for each journal. 
To improve the identification of hybrid journals, journals with an open access proportion above 95% were excluded.
This step allowed removing additional `r length(unique(oa_prop$issn_l))` fully open access journals.

Affiliation metadata about corresponding authors are crucial for the planning and evaluation of transformative agreements because they are considered responsible for arranging open access publication [@Schimmer_2015; @Geschuhn_2017; @Borrego_2020]. 
For this study, country and institutional affiliations were retrieved from OpenAlex.
However, because corresponding author affiliation was only available for 54% of the investigated articles in OpenAlex, this study focused on first authors and their affiliations instead; approximately 90% of the investigated articles had first author affiliation metadata in OpenAlex.
First authors typically contribute the most to a paper and are often considered being lead authors [@Larivi_re_2016]. 
Related studies also assumed first authors as a proxy to measure open access payments and the impact of transformative agreements [@Zhang_2022; @Shu_2023; @Haucap_2021].

@@

To assess the impact of transformative agreements on hybrid open access, participating institutions from the Transformative Agreement Data dump, which were crowd-sourced from the agreements and consortia that successfully negotiated an agreement, were matched with the first author affiliations recorded by OpenAlex using the ROR-ID.
Matching also considered the duration of agreements according to the ESAC registry.
Upon inspection, Transformative Agreement Data did not comprehensively cover associated institutions, such as university hospitals or institutes of large research organisations such as the Max Planck Society.
To improve the matching, Transformative Agreement Data were automatically enriched with associated organisations using OpenAlex's institution entity data.

The so compiled data set consists of `r hoaddata::jn_ind |> filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |> distinct(issn_l, cr_year, jn_all) |> pull(jn_all) |> sum()` articles published in `r hoaddata::jn_ind |> filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |> distinct(issn_l) |> nrow()` hybrid journals between 2018 and 2022 (see Figure \ref{fig:method_fig}B). 
Hybrid journals in transformative agreements represented `r cr_stats |> filter(src == "hoad") |> pull(n_articles) |> sum() / cr_stats |> pull(n_articles) |> sum() * 100`%  of the total global output over the same time period, according to Crossref, while fully open access journals recorded 
`r cr_stats |> filter(is.na(src) & oa_status == "gold") |> pull(n_articles) |> sum() / cr_stats |> pull(n_articles) |> sum() * 100`%.

```{r method_fig, fig.asp=0.55, fig.cap="Initial data characteristics. (A) Full open access journals included in transformative agreements by evidence source Directory of Open Access Journals (DOAJ), OpenAlex and the Bielefeld GOLD OA list. (B) Number of articles in Crossref by journal types. The blue bars show the overall article volume of hybrid journals in transformative agreements, which were initially included in the study, in comparision with full open access journals according to OpenAlex. The remainder represents subscription-based journals not covered by transformative agreements."}
(p | cr_stats_plot) + plot_annotation(tag_levels = 'A') +  plot_layout(widths = c(0.6, 0.4))
```
<!-- To reflect the increasing trend towards multiple and cross-country affiliations (Hottenrott et al., 2021; Larivière et al., 2015), articles were counted once per country. -->

@@

Throughout this mostly automated data gathering and analysis process, Tidyverse tools (Wickham et al., 2019) for the R programming language (R Core Team, 2020) were used. 
The resulting data are openly available through an R data package, hoaddata, version 0.2.91 
Following @Marwick_2018, hoaddata contains not only the datasets used in the data analysis. 
It also includes code used to compile the data by connecting it to the cloud-based Google Big Query data warehouse, where scholarly big data from Crossref, OpenAlex and Unpaywall were imported.
To increase computational reproducibility, data aggregation through hoaddata was automatically carried out using GitHub Actions, a continuous integration service.

@@

Between 2018 and 2022, `r n_hybrid_jns` out of `r n_hybrid_jns_all` hybrid journals in transformative agreements published at least one open access article under a Creative Commons license.
During this period, these hybrid journals provided open access to `r n_oa` of the `r n_total` articles, representing a five-year open access proportion of `r n_oa / n_total * 100`%.
Authors who could make use of transformative agreements at the time of publication contributed `r n_ta_oa` open access articles.

@@

Figure \ref{fig:results_overview}A shows a moderate growth in the proportion of open access articles in hybrid journals, comparing the overall open access uptake and the impact of transformative agreements on this trend.
Over the five-years period from 2018 to 2022, open access increased from `r oa_by_year |> filter(cr_year == "2018") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2018") |> pull(oa)`) to `r oa_by_year |> filter(cr_year == "2022") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2022") |> pull(oa)`). 
Simultaneously, the total article volume of the investigated journals increased from `r oa_by_year |> filter(cr_year == "2018") |> pull(articles)` in 2018 to `r oa_by_year |> filter(cr_year == "2022") |> pull(articles)` in 2022.

Figure \ref{fig:results_overview}B highlights that the majority of hybrid open access was made available through transformative agreements in 2021 and 2022, contributing `r oa_ta_plot_df |> filter(cr_year == "2022", type == "ta_oa") |> pull(value) / oa_by_year |> filter(cr_year == "2022") |> pull(oa) * 100`% of the total open access article volume in 2022.
However, there was also a notable increase in open access provision through other means, presumably publication fees being not invoiced through transformative agreements, which increased from `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> pull(value)`) in 2018 to  `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> pull(value)`). 
<!-- This suggests that publishers were able to gain equally from individual and institutional open access publishing options.-->
